Fast approximate Bayesian inference of HIV indicators using PCA adaptive Gauss-Hermite quadrature

### 2024-12-29 src/docs_paper-jtb/paper.Rmd
Format Section 5.4.1

### Point estimates \label{sec:point-estimates}

@@

### Exceedance probabilities\label{sec:exceedance}

#### Meeting the second 90

Ambitious targets for scaling up ART treatment have been developed by UNAIDS, with the goal of ending the AIDS epidemic by 2030 [@unaids2014target].
Meeting the 90-90-90 fast-track target requires that 90% of people living with HIV know their status, 90% of those are on ART, and 90% of those have a suppressed viral load.
Inferences from Naomi can be used to identify treatment gaps by calculating the probability that the second 90 target has been met, that is $\mathbb{P}(\alpha_i > 0.9^2 = 0.81)$ for each strata $i$.

(ref:exceedance-second90) The probability each strata has met the second 90 (ART coverage above 81\%) calculated using each inference method, as compared with NUTS. The root mean square error (RMSE) and mean absolute error (MAE) are displayed in the top left.

```{r exceedance-second90, fig.cap="(ref:exceedance-second90)"}
knitr::include_graphics("depends/exceedance-second90.png")
```

Strata probabilities of having met the second 90 target were more accurately estimated by GPCA-AGHQ than GEB (Figure \@ref(fig:exceedance-second90)).
Both GPCA-AGHQ and GEB had substantial error as compared to results from NUTS, however, particularly for girls and women.
This discrepancy in accuracy by sex may be caused by interactions between the household survey and ANC components of the model creating a more challenging posterior geometry.

#### Finding strata with high incidence {#one-inc}

Some HIV interventions are cost-effective only within high HIV incidence settings, typically defined as higher than 1% incidence per year.
Inferences from Naomi can be used to calculate the probability of a strata having high incidence by evaluating $\mathbb{P}(\lambda_i > 0.01)$.

(ref:exceedance-1inc) The probability each strata has high HIV incidence (above 1\% per year) calculated using each inference method, as compared with NUTS. The root mean square error (RMSE) and mean absolute error (MAE) are displayed in the top left.

```{r exceedance-1inc, fig.cap="(ref:exceedance-1inc)"}
knitr::include_graphics("depends/exceedance-1inc.png")
```

GPCA-AGHQ gave more accurate estimates of the probability that a strata has high HIV incidence than GEB (Figure \@ref(fig:exceedance-1inc)).
Again, both methods had significant error.
Unlike in Section \@ref(second90), there was little difference in error by sex.

>>>>

### Point estimates\label{sec:point-estimates}

```{r}
df_point <- readr::read_csv("depends/mean-sd.csv")

df_point_pct <- df_point %>%
  ungroup() %>%
  group_by(indicator, type) %>%
  summarise(
    rmse_diff = 100 * diff(rmse) / max(rmse),
    mae_diff = 100 * diff(mae) / max(mae)
  )

rmse_aghq_mean <- filter(df_point, method == "GPCA-AGHQ", indicator == "Posterior mean", type == "latent") %>% pull(rmse) %>% signif(2)
rmse_tmb_mean <- filter(df_point, method == "GEB", indicator == "Posterior mean", type == "latent") %>% pull(rmse) %>% signif(2)
rmse_diff_mean <- filter(df_point_pct, indicator == "Posterior mean", type == "latent") %>% pull(rmse_diff) %>% signif(2)

rmse_aghq_sd <- filter(df_point, method == "GPCA-AGHQ", indicator == "Posterior SD", type == "latent") %>% pull(rmse) %>% signif(2)
rmse_tmb_sd <- filter(df_point, method == "GEB", indicator == "Posterior SD", type == "latent") %>% pull(rmse) %>% signif(2)
rmse_diff_sd <- filter(df_point_pct, indicator == "Posterior SD", type == "latent") %>% pull(rmse_diff) %>% signif(0)
```

```{r mean-sd-alt-latent, fig.cap="The latent field posterior mean and posterior standard deviation point estimates from each inference method as compared with those from NUTS. The root mean square error (RMSE) and mean absolute error (MAE) are displayed in the top left. For the posterior mean and posterior standard deviation, GPCA-AGHQ reduced RMSE and MAE as compared with GEB."}
knitr::include_graphics("depends/mean-sd-alt-latent.png")
```

Latent field mean and standard deviation point estimates obtained from GPCA-AGHQ were more accurate than those from GEB (Figure \ref{fig:mean-sd-alt-latent}).
The root mean square error (RMSE) between posterior mean estimates from GPCA-AGHQ and NUTS (`r rmse_aghq_mean`) was `r abs(rmse_diff_mean)`% lower than that between GEB and NUTS (`r rmse_tmb_mean`).
For the posterior standard deviation estimates, there was a substantial `r abs(rmse_diff_sd)`% reduction in RMSE: from `r abs(rmse_tmb_sd)` (TMB) to `r abs(rmse_aghq_sd)` (PCA-AGHQ).
However, improvements in latent field estimate accuracy transferred to model outputs to a more limited and mixed extent (Supplementary Figures 12 and 13).

@@

#### Kolmogorov-Smirnov

```{r}
ks_summary <- readRDS("depends/ks-summary.rds")

ks_summary_summarise <- ks_summary %>%
  ungroup() %>%
  filter(type == "Latent") %>%
  summarise(
    tmb = sum(size * TMB) / sum(size),
    aghq = sum(size * aghq) / sum(size)
  )

ks_pct <- (ks_summary_summarise$tmb - ks_summary_summarise$aghq) / ks_summary_summarise$tmb
```

```{r ks-summary, fig.cap="The average Kolmogorov-Smirnov (KS) test statistic for each latent field parameter of the Naomi model. Vectors of parameters were grouped together. For points above the dashed line at zero, performance of GEB was better. For points below the dashed line, performance of GPCA-AGHQ was better. Most notably, for the latent field parameters ui_lambda_x the test statistic for GEB was substantially higher than for GPCA-AGHQ. This parameter, of length 32, corresponds to and plays a key role in the ART attendance component of the Naomi."}
knitr::include_graphics("depends/ks-summary.png")
```
